---
title: "A first multi-species size-spectrum model"
author: "Gustav Delius"
output: 
  html_notebook:
    toc: TRUE
---

We continue our investigation of size-spectrum dynamics in simple examples.
Where in the previous tutorial notebook we had studied the size-spectrum
dynamics of a single species in a fixed power-law background, here we will study
a species in a multi-species background, where the background species interact
with the foreground species and vice versa.

Again in this tutorial we are not yet interested in a realistic model, but in
a model that helps us to get a feel for the multispecies effects that we can
expect. We start looking at the real world only in tomorrow's notebook.

Start RStudio 
and load the following packages:
```{r message=FALSE}
library(mizerExperimental)
library(tidyverse)
library(plotly)
library(ggplot2)
```

# A trait-based multi-species model

First we will create a generic model consisting of several interacting species.
We use another wrapper function `newTraitParams()` that sets up a MizerParams
object choosing a lot of defaults for us.
```{r}
params_background <- newTraitParams(
    lambda = 2.12,      # Exponent of community spectrum
    no_sp = 16,         # Number of species
    min_w_inf = 1,      # Asymptotic size of smallest species
    max_w_inf = 10^6,   # Asymptotic size of largest species
    min_w = 10^-5,      # Egg size
    min_w_mat = 0.1,    # Maturity size of smallest species
    min_w_pp = 10^-12,  # Size of smallest plankton
    ext_mort_prop = 0.2 # Proportion of mortality that is not due to predation
    )

plotSpectra(params_background, power = 2, total = TRUE)
```

We only had to specify a small number of parameters and the function makes
default choices for all the others.

Note how the total community, aggregated over all species and the plankton,
follows a Sheldon spectrum with the Sheldon exponent `lambda` of our choice.

These species will form our background, modelling all the species about which we
do not have much information and don't really care about. We tell mizer about
this with
```{r}
params_background <- markBackground(params_background)
```
This will for example have the effect that these species are plotted in grey.
We also don't want to fish them, so we set their catchability to zero.
```{r}
gear_params(params_background)$catchability <- 0
```



# Adding a foreground species

Let's look at hake. Let's assume we know the following parameters for hake:
```{r}
species_params <- tribble(
    ~species, ~w_inf, ~w_mat, ~beta, ~sigma, ~a,      ~b,    ~k_vb, ~l50, ~l25,
    "Hake",   10470,  250,    11,    1.1,    0.00667, 3.035, 0.178, 16.6, 16
)
species_params$sel_func <- "sigmoid_length"
```
Here `w_inf` is the asymptotic size, `w_mat` the maturity size, `beta` and
`sigma` the parameters of the log-normal predation kernel, `a` and `b` the
parameter in the length to weight conversion $w = a\, l^b$, `k_vb` the von
Bertalanffy growth curve parameter and `l50` and `l25` the parameters of the
`sigmoid_length()` selectivity function.

In practice you would probably not type them into R as above but create a 
spreadsheet with the parameters and then read that in with `read_csv()`.

We can now add Hake to our existing model with the `addSpecies()` function.
```{r}
params <- addSpecies(params_background,
                     species_params,
                     initial_effort = c(knife_edge_gear = 1)) %>% 
    steady()
plotSpectra(params, power = 2, total = TRUE)
```

# Increasing abundance of foreground species
Initially this has added hake with a low abundance, so that there is little
effect of the presence of the hake on the background species. We can now
slowly increase the abundance of the hake, making sure we always run to steady
state.

First we double the abundance of the foreground species with respect to the
background species.
```{r}
params <- rescaleAbundance(params, factor = 2) %>% 
    steady()
plotSpectra(params, power = 2, total = TRUE)
```

Now we double it again, which we can do simply by running the previous R 
chunk again.

After doing this 4 times we obtain
```{r echo=FALSE}
plotSpectra(params, power = 2, total = TRUE)
```

The multi-species effects are now clearly visible.

# Exploring with the tuning gadget
Instead of exploring further by hand, we now switch to a convenient shiny
gadget, which provides sliders for changing parameters and a set of plots that
update in real time to reflect the effects of the parameter changes. I will
demonstrate this on the projector.

```{r eval=FALSE}
params_new <- tuneParams(params)
```
